JD notes on effects

### 2022-07-07 manuscript/body.tex
More progress

% Use "Eq" instead of "Equation" for equation citations.
% Best to use a command that combines the name and the reference
\section{Introduction}

@@

\item \textbf{Input variables}: Refers to the observed (or scientific) variables underlying an inference or exploration. For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
\item \textbf{Focal and non-focal predictors}: The focal predictor refers to the input variable we are interested in. For multivariate models, all other variables other than focal predictors are non-focal predictors. For example, consider model described by Equation~\ref{eq:simple_inter_higher_no_interaction}. Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.

@@

\item \textbf{Model variables}: Refer to the X-variables which go into the regression model; and represent columns in the model matrix. Each input variable may correspond to one or more model variable. In particular, variables with more than two categories (for instance religion in our previous example), or input variables modeled with interactions, a spline or polynomial interactions, will correspond to more than one model variable. For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} have $4$ model variables -- $x_1, x_2, x_3$ plus intercept term and $5$ model variables -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables). Also referred to as center point. We refer to the prediction approach that uses model center as mean-based. The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}B. In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for the center point. If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see Fig~\ref{fig:pred_cubic_plots}.

@@

\item \textbf{Marginal effects:} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable. This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables. Consider models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher}. The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$. On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$. In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$. In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope multi-dimensional surface with response to the surface of interest \cite{leeper2017interpreting}. 

@@

We could also have interactions between focal and non-focal predictors. To illustrate, we consider simple input variables but the logic extends to more complex ones like MPVs. Consider model described by Equation~\ref{eq:simple_inter_higher}, used to demonstrate how to handle interactions in non-focal predictors. Here, we use it to demonstrate how to handle interactions between the focal, $x_2$, and non-focal, $x_3$, predictors. In this case, the non-interacting predictors are treated as non-focal model variables, with appropriate choice of reference points. However, for the interacting model variables, we first choose representative values for the focal predictor and then some particular values of the interacting non-focal predictors in a similar way we choose the values of the focal predictor or use predetermined values. In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting non-focal predictor, $x_3$, respectively. The central estimate on linear predictor scale is given by
%
\begin{align*}
\eta^\star(x_{2i}, x_{3j}, {\bar{x}^\star_1}) = \beta_0 + \beta_1 \bar{x}^\star_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
\end{align*}
%

@@

Let $\bxo$ be a centered model matrix previously defined, and let $\ba$ be an anchor matrix, with the same dimensions and entries in all non-focal predictors as $\bxo$. Let $\baf$ be the column of $\ba$ corresponding to focal predictor(s) defined in $\bxo$. Any appropriate values can be chosen for $\baf$ but for model center (center-anchored), we use $\bafc$ which is the mean of the focal predictor(s). Thus 
%
\begin{align}\label{eq:centered_variance}
\boldsymbol\sigma_f^2 = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top).
\end{align}
%
We can see that $\forall ~\bxof=\baf$, $\bxo - \baf = \boldsymbol 0$, hence $\boldsymbol\sigma_f^2 = \boldsymbol{0}$. Similarly, for all values of $\bxof$ close to anchor point $\bafc$, the term $(\bxo - \bafc)$ and $\boldsymbol\sigma_f^2$ goes to $\boldsymbol 0$ and $\boldsymbol\sigma_f^2 = \boldsymbol 0$ if $\bxof=\bafc$. This means that $\boldsymbol\sigma_f^2$ close to the anchor point are smaller than those away from the anchor point; and results to confidence intervals which are narrower around the anchor or crosses at the anchor point for simple models. This is the effect of the particular focal input variable. By setting $\baf = \boldsymbol 0$, we get the variances for the conventional CIs in Equation~\ref{eq:conventional_variance}.

@@

An alternative approach to choosing a reference point is to make predictions over all observations of the non-focal input variables (members of the population) \citep{hanmer2013behind}. The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in Equation~\ref{eq:eta_mean}. More specifically:
\begin{itemize}
\item compute linear predictor associated with whole sample of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
\item compute linear predictor associated with the chosen values focal input variable, $\eta_{jf} = \sum{\beta_f x_{jf}}$
\item for every value of the focal linear predictor, $\eta_{jf}$, compute
%
\begin{align}\label{eq:pop_eta} 
\eta_j^\star(\eta_{jf}, \nset{\eta})  &= \eta_{jf} + \nset{\eta} \nonumber \\
&= \eta_j^\star(x_f, \nset{x}).
\end{align}
\end{itemize}
%

Once Equation~\ref{eq:pop_eta} is computed, we back-transform the estimates to the original scale and average over the levels of the focal predictors, $j$:
%
\begin{align}\label{eq:pop_response} 
\hat{y}_f  &= \textrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x})\right).
\end{align}
%

@@

For models with random effects components, we make further adjustment to correct for bias induced by the random effects terms. In whole-sample approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to Equation~\ref{eq:pop_eta}. In particular
%
\begin{align}\label{eq:pop_eta_re} 
\tau &= \bZ b \nonumber \\
\eta_j^\star(x_f, \nset{x}, \tau)  &= \eta_j^\star(x_f, \nset{x}) + \tau
\end{align}
where $\bZ$ and $b$ are the design matrix and a vector of random effects, respectively.

@@

In the whole-sample-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see Equation~\ref{eq:pop_response}, so we do not need to worry about the nonlinear averaging. In other words, the averaging is no longer on the link scale and is likely to be bounded by the original data scale. In simple linear models without interactions, averaging on the link scale is identical to averaging on the response, so both approaches yield similar results. However, picking a single value, e.g., the mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic, or not contained in or representative of the population. In addition, the mean-based approach fails to use every value of non-focal predictors hence not utilizing the full potential of the information contained in the data. This may limit the inferences we can make about the entire population. In general, the mean-based approach provides the predictions of an average case, whereas the whole-sample-based approach summarizes the predictions over the entire population. In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population. This might not be a problem in the whole-sample-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal input variable.

@@

We now illustrate the construction of prediction and effect plots and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias. In addition, we demonstrate that whole-sample-based approach can be used for bias correction.

@@

The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively. However, there are a number of choices one can make when generating predictions -- for example, in the presence of interactions, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}). For simple OLS models (with no interactions), the two approaches give the same predictions. However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal input variable:
%
\begin{align}
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}.
\end{align}
%

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
%
\begin{figure}
\begin{center}
\includegraphics{justify_preds.inter.pdf}
\end{center}
\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for regression of $x_1$ on $y$.} The horizontal dashed lines are the mean of the predicted and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively. The grey points are the binned simulated $y$. The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their mean and the true predictions based on the simulation parameters. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated value in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. Similarly, the predictions based on the three approaches match the true predictions. B: Even with the simple interaction between non-focal input variables, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}
%
In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A. However, in the presence of interactions, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates start to differ. In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B. To generate Fig~\ref{fig:justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the predictor variables corresponds to $x_2$ and $x_3$. On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$). In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses predictor variables to compute $\over{x_2x_3}$. In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the predictor variables.

One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects. However, currently, the two packages do not provide a straightforward way do achieve this. To illustrate this, we generate predictions of $x_2$ from Equation~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence intervals, as shown in Fig~\ref{fig:justify_ci_plots}. 
%
\begin{figure}
\begin{center}
\includegraphics{justify_preds.isolate.pdf}
\end{center}
\caption{{\bf The prediction and effect styled plots at $95\%$ confidence intervals.} The description of horizontal, vertical and trend lines remain the same as above. The wider dashed curves corresponds to the conventional confidence intervals around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower curves crossing at the mean of the focal input variable, i.e., the center point, correspond to the effect of the focal input variable from \pkg{varpred}. For simple OLS models, the effect-styled confidence bands crosses at the center point (an anchor). The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables. On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal input variable.}
\label{fig:justify_ci_plots}
\end{figure}
%
For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}. In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal input variable we are more certain about, i.e., the center point. For simple OLS models, the point where the confidence intervals curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate choice). Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor. The lines inside (effects) are a little different; they are a good way to visualize the effect of the focal predictor. Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.

\jd{Why do we need this in the definitions section?:} In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the center (mean) of the focal input variable with \ldots.
%
\begin{figure}
\begin{center}
\includegraphics{justify_anchors.combined.pdf}
\end{center}
\caption{{\bf A comparison of zero- and center-anchored effects.} \jd{This explanation is not currently correct, but do we really need to explain that the zero anchor means that the anchor is at zero? Also, the real difference between anchors is for multi-parameter predictors? Do we have an example of that?} The center-anchored effects are based on the mean of the focal predictor while the zero-anchored effects are based on the minimum of the focal predictor. As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).} 
\label{fig:justify_anchors}
\end{figure}
%

@@

To illustrate the distinction between the two approaches, we simulate a multivariate model described in \nameref{S1_Appendix}. We also use the same model to illustrate how predictions and effects differ in the presence of higher order interactions in focal or non-focal input variables. We then generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial focal input variable); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable). In both cases the, the center point was used as the anchor for the effects, as shown in Fig~\ref{fig:pred_cubic_plots}.

@@

Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions. To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that age has a very small effect size in comparison to wealth index, and compared their effects on the predicted probability of improved water quality as shown in Fig~\ref{fig:pred_bin_plots}.

@@

If there was no effect of nonlinear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in Fig~\ref{fig:pred_cubic_plots}B. One possible reason for the variations we see in Fig~\ref{fig:pred_bin_plots} is the nonlinear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale. For example, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the center point.

In Fig~\ref{fig:pred_bin_prediction_effects_plots} below, we zoom into Fig~\ref{fig:pred_bin_plots}{A} and compare the prediction and effect styled plots.

\begin{figure}
\begin{center}
\includegraphics{glm_two_predictor_preds_effects.ggp.pdf}
\end{center}
\caption{{\bf Prediction and effect styled plots and bias correction.} The central curves for predictions and effects are the same in the respective case. These patterns are  consistent with the ones in Fig~\ref{fig:pred_bin_plots}A with respect to the \emph{truth}. Further, the prediction and effect CIs exhibit patterns similar to the ones we previously described and generalizes in a similar way to the ones in simple linear models.
}
\label{fig:pred_bin_prediction_effects_plots}
\end{figure}

>>>>

% Use "Eq" instead of "Equation" for equation citations. 
%% BC: use \EREF{} for equations and \FREF{} for figures
% Best to use a command that combines the name and the reference
\section{Introduction}

@@

\item \textbf{Input variables}: Refers to the observed (or scientific) variables underlying an inference or exploration. For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} both have $3$ input variables -- $x_1, x_2, x_3$.
\item \textbf{Focal and non-focal predictors}: The focal predictor refers to the input variable we are interested in. For multivariate models, all other variables other than focal predictors are non-focal predictors. For example, consider model described by \EREF{simple_inter_higher_no_interaction}. Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.

@@

\item \textbf{Model variables}: Refer to the X-variables which go into the regression model; and represent columns in the model matrix. Each input variable may correspond to one or more model variable. In particular, variables with more than two categories (for instance religion in our previous example), or input variables modeled with interactions, a spline or polynomial interactions, will correspond to more than one model variable. For example, the regression models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher} have $4$ model variables -- $x_1, x_2, x_3$ plus intercept term and $5$ model variables -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables). Also referred to as center point. We refer to the prediction approach that uses model center as mean-based. The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in \FREF{justify_plots}B. In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for the center point. If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the focal predictor, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see \FREF{pred_cubic_plots}.

@@

\item \textbf{Marginal effects:} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable. This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables. Consider models described by \EREF{simple_inter_higher_no_interaction} and \EREF{simple_inter_higher}. The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$. On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$. In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$. In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope multi-dimensional surface with response to the surface of interest \cite{leeper2017interpreting}. 

@@

We could also have interactions between focal and non-focal predictors. To illustrate, we consider simple input variables but the logic extends to more complex ones like MPVs. Consider model described by \EREF{simple_inter_higher}, used to demonstrate how to handle interactions in non-focal predictors. Here, we use it to demonstrate how to handle interactions between the focal, $x_2$, and non-focal, $x_3$, predictors. In this case, the non-interacting predictors are treated as non-focal model variables, with appropriate choice of reference points. However, for the interacting model variables, we first choose representative values for the focal predictor and then some particular values of the interacting non-focal predictors in a similar way we choose the values of the focal predictor or use predetermined values. In our example, suppose we pick $i$ and $j$ unique values of the focal predictor, $x_2$ and interacting non-focal predictor, $x_3$, respectively. The central estimate on linear predictor scale is given by
%
\begin{align*}
\eta^\star(x_{2i}, x_{3j}, {\bar{x}^\star_1}) = \beta_0 + \beta_1 \bar{x}^\star_1 + \beta_2x_{2i} + \beta_3x_{3j} + \beta_{23}x_{2i}x_{3j}.
\end{align*}
%

@@

Let $\bxo$ be a centered model matrix previously defined, and let $\ba$ be an anchor matrix, with the same dimensions and entries in all non-focal predictors as $\bxo$. Let $\baf$ be the column of $\ba$ corresponding to focal predictor(s) defined in $\bxo$. Any appropriate values can be chosen for $\baf$ but for model center (center-anchored), we use $\bafc$ which is the mean of the focal predictor(s). Thus 
%
\begin{align}\label{eq:centered_variance}
\boldsymbol\sigma_f^2 = \textrm{Diag}((\bxo - \bafc) \boldsymbol{\Sigma} (\bxo - \bafc)^\top).
\end{align}
%
We can see that $\forall ~\bxof=\baf$, $\bxo - \baf = \boldsymbol 0$, hence $\boldsymbol\sigma_f^2 = \boldsymbol{0}$. Similarly, for all values of $\bxof$ close to anchor point $\bafc$, the term $(\bxo - \bafc)$ and $\boldsymbol\sigma_f^2$ goes to $\boldsymbol 0$ and $\boldsymbol\sigma_f^2 = \boldsymbol 0$ if $\bxof=\bafc$. This means that $\boldsymbol\sigma_f^2$ close to the anchor point are smaller than those away from the anchor point; and results to confidence intervals which are narrower around the anchor or crosses at the anchor point for simple models. This is the effect of the particular focal input variable. By setting $\baf = \boldsymbol 0$, we get the variances for the conventional CIs in \EREF{conventional_variance}.

@@

An alternative approach to choosing a reference point is to make predictions over all observations of the non-focal input variables (members of the population) \citep{hanmer2013behind}. The nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population to a one-dimensional distribution of $\eta^\star(x_f, \nset{x})$, which is a function of the chosen values of the focal predictor and the whole sample of non-focal predictors, as opposed to the definition in \EREF{eta_mean}. More specifically:
\begin{itemize}
\item compute linear predictor associated with whole sample of the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
\item compute linear predictor associated with the chosen values focal input variable, $\eta_{jf} = \sum{\beta_f x_{jf}}$
\item for every value of the focal linear predictor, $\eta_{jf}$, compute
%
\begin{align}\label{eq:pop_eta} 
\eta_j^\star(\eta_{jf}, \nset{\eta})  &= \eta_{jf} + \nset{\eta} \nonumber \\
&= \eta_j^\star(x_f, \nset{x}).
\end{align}
\end{itemize}
%

Once \EREF{pop_eta} is computed, we back-transform the estimates to the original scale and average over the levels of the focal predictors, $j$:
%
\begin{align}\label{eq:pop_response} 
\hat{y}_f  &= \textrm{mean} ~ g^{-1} \left(\eta_j^\star(x_f, \nset{x})\right).
\end{align}
%

@@

For models with random effects components, we make further adjustment to correct for bias induced by the random effects terms. In whole-sample approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to \EREF{pop_eta}. In particular
%
\begin{align}\label{eq:pop_eta_re} 
\tau &= \bZ b \nonumber \\
\eta_j^\star(x_f, \nset{x}, \tau)  &= \eta_j^\star(x_f, \nset{x}) + \tau
\end{align}
where $\bZ$ and $b$ are the design matrix and a vector of random effects, respectively.

@@

In the whole-sample-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see \EREF{pop_response}, so we do not need to worry about the nonlinear averaging. In other words, the averaging is no longer on the link scale and is likely to be bounded by the original data scale. In simple linear models without interactions, averaging on the link scale is identical to averaging on the response, so both approaches yield similar results. However, picking a single value, e.g., the mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic, or not contained in or representative of the population. In addition, the mean-based approach fails to use every value of non-focal predictors hence not utilizing the full potential of the information contained in the data. This may limit the inferences we can make about the entire population. In general, the mean-based approach provides the predictions of an average case, whereas the whole-sample-based approach summarizes the predictions over the entire population. In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population. This might not be a problem in the whole-sample-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal input variable.

@@

We now illustrate the construction of outcome plots, compare with other packages and also demonstrate that the mean-based and whole-sample-based approaches produce different results in models with nonlinear link functions and/or additional source(s) of potential bias. In addition, we demonstrate that whole-sample-based approach can be used for bias correction.

@@

The most commonly used \proglang{R} software packages for outcome plots (\pkg{emmeans} and \pkg{effects}), by default, use the average of input variables and ``sum-to-zero'' as the reference point approach for continuous and categorical input variables, respectively. However, there are a number of choices one can make when constructing outcome plots -- for example, in the presence of interactions, averaging the interactions (averaging the product of model variables) versus taking the product of the averages of the input variables (the default for \pkg{emmeans} and \pkg{effects}). For simple OLS models (with no interactions), the two approaches give the estimates. However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal input variable:
%
\begin{align}
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}.
\end{align}
%
We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare estimates from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' central estimates (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in \FREF{justify_plots}.
%
\begin{figure}
\begin{center}
\includegraphics{justify_preds.inter.pdf}
\end{center}
\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} central estimates for regression of $x_1$ on $y$ (outcome plots).} The horizontal dashed lines are the mean of the central estimates and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively. The grey points are the binned simulated $y$. The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their mean and the true predictions based on the simulation parameters -- central estimate. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated one in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. Similarly, the central estimate based on the three approaches match the truth. B: Even with the simple interaction between non-focal predictors, we see differences between predicted mean, $\bar{\hat{y}}$ and simulated mean , $\bar{y}$. Similarly, there are differences between central estimates and the truth, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}
%
In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, \FREF{justify_plots}A. However, in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}, the estimates start to differ. In particular, \pkg{emmeans} and \pkg{effects} have similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B. To generate \FREF{justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the model variables corresponding to $x_2$ and $x_3$. On other hand, to generate \FREF{justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction model variable ($x_2x_3$). In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses model variables to compute $\over{x_2x_3}$. In the case of \FREF{justify_plots}A, averaging the input variables is equivalent to averaging the model variables.

\BC{Got to this point \dots}

One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effect plots. However, as previously mentioned, the two packages do not provide a straightforward way do achieve this. To illustrate this, we consider \EREF{simple_inter_higher_no_interaction} and use $x_2$ as the focal predictor, and generate the outcome plot shown in \FREF{justify_ci_plots}. 
%
\begin{figure}
\begin{center}
\includegraphics{justify_preds.isolate.pdf}
\end{center}
\caption{{\bf The prediction and effect styled plots at $95\%$ confidence intervals.} The description of horizontal, vertical and trend lines remain the same as above. The wider dashed curves corresponds to the conventional confidence intervals around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower curves crossing at the mean of the focal input variable, i.e., the center point, correspond to the effect of the focal input variable from \pkg{varpred}. For simple OLS models, the effect-styled confidence bands crosses at the center point (an anchor). The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables. On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal input variable.}
\label{fig:justify_ci_plots}
\end{figure}
%
For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}. In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal input variable we are more certain about, i.e., the center point. For simple OLS models, the point where the confidence intervals curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate choice). Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor. The lines inside (effects) are a little different; they are a good way to visualize the effect of the focal predictor. Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.

\jd{Why do we need this in the definitions section?:} In \FREF{justify_anchors}, we compare effects plots anchored at the center (mean) of the focal input variable with \ldots.
%
\begin{figure}
\begin{center}
\includegraphics{justify_anchors.combined.pdf}
\end{center}
\caption{{\bf A comparison of zero- and center-anchored effects.} \jd{This explanation is not currently correct, but do we really need to explain that the zero anchor means that the anchor is at zero? Also, the real difference between anchors is for multi-parameter predictors? Do we have an example of that?} The center-anchored effects are based on the mean of the focal predictor while the zero-anchored effects are based on the minimum of the focal predictor. As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).} 
\label{fig:justify_anchors}
\end{figure}
%

@@

To illustrate the distinction between the two approaches, we simulate a multivariate model described in \nameref{S1_Appendix}. We also use the same model to illustrate how predictions and effects differ in the presence of higher order interactions in focal or non-focal input variables. We then generated two sets of mean-based predictions and effects: 1) one with \code{age} as the focal input variable (a cubic polynomial focal input variable); and 2) one with \code{Wealthindex} as the focal in put variable (linear focal input variable). In both cases the, the center point was used as the anchor for the effects, as shown in \FREF{pred_cubic_plots}.

@@

Mean-based and whole-sample-based approaches can produce very different results in models with nonlinear link functions with additional sources of potential bias such as additional non-focal input variable(s), random effects or complex interactions. To demonstrate this, we considered a two predictor binary outcome simulation described in \nameref{S2_Appendix}, such that age has a very small effect size in comparison to wealth index, and compared their effects on the predicted probability of improved water quality as shown in \FREF{pred_bin_plots}.

@@

If there was no effect of nonlinear averaging, then we would expect the average observed proportion and the average predictions to intersect at the center point as we see in \FREF{pred_cubic_plots}B. One possible reason for the variations we see in \FREF{pred_bin_plots} is the nonlinear averaging; since both observed status and predicted probabilities are averaged on the response scale as opposed to link scale. For example, if the range of values are bigger than $0.5$ (seemingly the case here), then we would expected the averages to be slightly higher than what we would expect at the center point.

In \FREF{pred_bin_prediction_effects_plots} below, we zoom into \FREF{pred_bin_plots}A and compare the prediction and effect styled plots.

\begin{figure}
\begin{center}
\includegraphics{glm_two_predictor_preds_effects.ggp.pdf}
\end{center}
\caption{{\bf Prediction and effect styled plots and bias correction.} The central curves for predictions and effects are the same in the respective case. These patterns are  consistent with the ones in \FREF{pred_bin_plots}A with respect to the \emph{truth}. Further, the prediction and effect CIs exhibit patterns similar to the ones we previously described and generalizes in a similar way to the ones in simple linear models.
}
\label{fig:pred_bin_prediction_effects_plots}
\end{figure}
